Supporting Information:\\Black-swan events in animal populations

### 2016-06-24 text/ms.tex
More copyediting

Words in abstract: 142

@@

Number of references: todo

@@

Black swans are improbable events that nonetheless
occur---often with profound consequences. While physical extremes, such as
monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes. Here we estimate the frequency and
direction of black-swan events in \NPops\ animal populations, finding strong
evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
Black-swan events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%),
manifest primarily as
population die-offs and crashes (\percBSDown \%) rather than unexpected increases,
and tend to be driven by extrinsic perturbations
such as climate, severe winters, or disease.
Thus, ignoring black-swan events will
underestimate extinction risk.
% Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes.
Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to ecological surprises.

@@

Whereas there are many reports of black-swan events,
only a flexible comparative approach consistently applied 
to a large number of timeseries can yield insights into the frequency, strength,
and correlates of black-swan events. 
We are unaware of such a comparative analysis.
Previous comparative analyses fitting
heavy-tailed distributions
(distributions with higher probabilities of extreme events
than the normal distribution)
to timeseries
have not accounted for an underlying population dynamics model
\citep{keitt1998, allen2001, halley2002, segura2013}.
Alternatively, most examples of population die offs come from identifying
sudden changes in abundance that exceed a threshold of decline
\citep[e.g.][]{young1994, gerber2001, reed2003, fey2015}.
But this approach gives no indication of relative event frequency
and again does not distinguish events from
expected dynamics.
Methodologically, the generalized extreme value (GEV) distribution represents
the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for physical ecological variables
\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
the GEV requires maxima per time block
or peaks beyond a chosen threshold,
and hence requires longer timeseries then typically available
for wild populations.
Furthermore, the GEV approach cannot be easily integrated into population dynamics models.

@@

We fit population dynamics models in which the process noise is
drawn from a Student-$t$ distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the degree to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
Lower values of \(\nu\) result in heavy-tailed distributions.
% For example, at \(\nu = 2\), an event
% more than 4 standard deviations below the mean is expected
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution
% more than four standard deviations below the mean is almost 70,000 times higher
% than the probability of drawing such a value from a normal distribution
% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

@@

This could represent the effect of unmodeled spatial heterogeneity,
nonstationary environmental processes, or
varying responses to temperature among individuals, for example.

@@

We selected abundance time series from the Global Population Dynamics
Database \citep{gpdd2010} (GPDD), which contains nearly 5,000 time series of
abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
filtered the data (Supporting Information) to remove populations from less
reliable data sources, and those without sufficient data for our models, and
then interpolated remaining missing values (sensu \citet{brook2006a}).
Our interpolation affected only \(\sim\)
\interpPointsPerc \% of the final data points (Table~S1) and none of the data
points that were later considered black-swan events. Our final dataset
contained \NPops\ populations across \NOrders\ taxonomic orders and seven
taxonomic classes, with a median of \medianTimeSteps\ time steps (range
\minTimeSteps--\maxTimeSteps) (Table~S1).

@@

\subsubsection{Density independence} We fit a simplified version of the
Gompertz model in which the density dependence parameter \(b\) was fixed at
\(1\) (density independent). This is equivalent to fitting a random walk model
(with drift) to the \(\ln\) abundances or assuming the growth rates are drawn
from a stationary distribution. The model was as follows:
\begin{align}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align}
We fit this model for three reasons: (1) it is computationally simpler and so
provides a check that our more complicated full Gompertz model was obtaining
reasonable estimates of \(\nu\), (2) it provides a test of whether density
dependence was systematically affecting our perception of heavy tails, (3) it
matches how some previous authors have modelled heavy tails without accounting
for density dependence \citep{segura2013}.

\subsubsection{Observation error} Observation error can bias parameter
estimates \citep{knape2012} and is known to affect the ability to detect extreme
events \citep{ward2007}. In our main analysis, we fit a model that ignored
observation error. One way to account for observation error would be to fit a
full state-space model that simultaneously estimates the magnitude of process
noise and observation error. However, simultaneously estimating observation and
process noise is a challenging problem (e.g.\ because the observation and
process noise parameters tend to negatively covary in model fitting) and is
known to sometimes result in identifiability issues with the Gompertz
population model \citep{knape2008}. Furthermore, our model was applied to
hundreds of time series, often of short length (as few as \minTimeSteps\ time
steps) and our model estimates an additional parameter---the shape of the
process deviation tails---potentially making identifiability and computational
issues even greater. Therefore, we considered a version of the base Gompertz
model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
represents the standard deviation of observation error (on a log scale), and
\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
\begin{align}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
\end{align}
where \(r_\mathrm{max}\) represents the theoretical maximum population growth
rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
carrying capacity and, as before, \(x_t\) represents the \(\ln\) transformed
abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
population growth rate with increases in abundance. In contrast, the Gompertz
model assumes a linear decrease in population growth rate with increases in
\textit{ln} abundance.

@@

\textit{TODO: Should the following 2 paragraphs go in the results?}
We found that we could consistently and precisely recover median posterior
estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
200\)) (Fig.~S\ref{fig:sim-nu} upper panels). At smaller samples we could
usually qualitatively distinguish heavy from not-heavy tails, but the model
tended to underestimate how heavy the tails were. At the same time, at smaller
sample sizes, the model tended to overestimate how large the scale parameter
was (Fig.~S\ref{fig:sim-nu} lower panels).

@@

\subsection{Skewed Student-t forecasts} To evaluate the apparent skewness
of heavy-tailed process noise we fit Gompertz models with skewed Student-t
distributed process noise \citep{fernandez1998}. This distribution adds one
parameter to the Student-$t$ distribution, \(\gamma\), which controls the
skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
< \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly
informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We
aggregated 5,000 randomly selected posterior samples from the \(\gamma\)
parameter of each model at three levels of evidence for heavy-tails:
\emph{heavy tailed}, median \(\nu < 10\); \emph{slightly heavy tailed}, \(10 \le
\mathrm{median}\, \nu \le 70\); and \emph{normal tailed}, median \(\nu > 70\)
(Fig.~4).

@@

Given our results, we suggest modelers should routinely consider representing population dynamic
process noise with a heavy-tailed, and possibly downward-skewed, distribution,
especially when making forecasts to evaluate risk.
But what heavy-tailed distribution should one pick
and how heavy tailed should it be?
% nearly any reason for why a population modeler has incorrectly specified the exact and full process generating the abundance times could generate a heavy tailed process noise.
This is an open research question.
The exact shape of the tails has proven important
in the field of dispersal biology \citep{kot1996, clark1999}.
Answers for the field of population dynamics
might come from simulation analyses exploring
the implications of different tail shapes and
from comparing the relative fit of heavy-tailed
population models to timeseries
with suspected mass mortality events
\citep[e.g.~the $t$ distribution compared to a mixture
of two normal distributions;][]{ward2007}.
Nevertheless, our proposed $t$ (or the skew-$t$) distribution
is a simple and elegant solution that
can be easily substituted for the normal distribution
Furthermore, if the degrees of freedom parameter, $\nu$,
is fixed
it requires estimating no more parameters
than the normal distribution.
Barring a better estimate,
we suggest using $\nu = 4$; and
a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a)
if a modeler wishes to allow for downward heavy-tailed events.
These are midrange values from the populations identified as heavy tailed here.
Above $\nu  =  10$ the $t$ distribution closely resembles
the normal distribution,
and $\nu  = 4$ allows for an event
three standard deviations below the mean to happen
once every 50 years instead of once every 740 years.
Adopting heavy-tailed process noise by default follows the intuition
that our models are usually reasonable but occasionally very wrong.

@@

Fig.~1: Illustration of population dynamic models that allow for heavy
tails. (a to b) Probability density for the Student-t
distribution with scale parameter of 1 and different values of \(\nu\). Small
values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
distribution approaches the normal distribution. (c to e)
Simulated population dynamics from a Gompertz model with process noise drawn
from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
panels c and d represent jumps with less than a 1 in 1000
chance of occurring in a normal distribution. (f to h)
Estimates of \(\nu\) from models fit to the times series in c to
e. Shown are posterior samples (histograms), median and interquartile
range of the posterior (IQR, dots and line segments), and the exponential prior
on \(\nu\) (dashed lines). Colour shading behind f to h
illustrates the approximate region of heavy tails.

@@

Fig.~2: Estimates of population dynamic heavy-tailedness for
populations of birds, mammals, insects, and fishes. (a to d)
Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
points and line segments represent posterior medians and 50\% / 90\% credible
intervals for individual populations. Inset plots show probability that \(\nu
< 10\) for populations arranged by taxonomic order and sorted by decreasing
mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
panel a are omitted for space. Red to yellow points highlight populations
with a high to moderately high probability of heavy-tailed dynamics.

@@

Fig.~3: Standardized probabilities and covariates of heavy-tailed
dynamics. (a) Taxonomic-order-level posterior densities of Pr(\(\nu
< 10\)) after accounting for time-series length in a hierarchical model. Dotted
vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the
prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
green: insects, purple: birds, and red: mammals). Estimates are standardized at
the geometric mean of time series length across all the data (approximately 27
time steps) and shown for orders with \(\ge 5\) populations. (b)
Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
panels, short vertical line segments within the density polygons indicate
medians.

@@

Fig.~4: Heavy-tailed process noise tends to be downward skewed and
ignoring this can underestimate risk. (a) Illustrations of Student-t
distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
(b) Posterior density of the skewness parameters aggregated across
populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly
heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations
(\(\hat{\nu} > 70\)). Approximate mid-values from b are illustrated in
panel a. (c) Example time series of heavy-tailed populations with
different levels of skewness. Red dots highlight likely heavy-tailed events.
Forecasts (grey regions) show median (solid lines) and lower 99\% credible
intervals (dotted lines) of abundance. Black and red lines indicate forecasts
from Gompertz models with lognormal and skew-$t$ process noise, respectively.

>>>>

Words in abstract: 145

@@

Number of references: 55

@@

Black swans are improbable events that nonetheless
occur---often with profound consequences. While physical extremes, such as
monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological population numbers
buffer or suffer from such extremes. Here we estimate the frequency and
direction of black-swan events in \NPops\ animal populations, finding strong
evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations.
Black-swan events occur most frequently for birds (\birdPH \%), mammals
(\mammalsPH \%), and insects (\insectsPH \%),
manifest primarily as
population die-offs and crashes (\percBSDown \%) rather than unexpected increases,
and tend to be driven by extrinsic perturbations
such as climate, severe winters, or disease.
Thus, ignoring black-swan events will
underestimate extinction risk.
Our results demonstrate the importance of both
modeling heavy-tailed downward events in populations, and developing
conservation strategies that are robust to ecological surprises.
% Ignoring black-swan events leads to a \crashUnderMedian-fold (50\% interval \crashUnderRange) underestimate in the magnitude of population crashes.

@@

Whereas there are many reports of black-swan events,
only a flexible comparative approach consistently applied 
to a large number of time series can yield insights into the frequency, strength,
and correlates of black-swan events. 
We are unaware of such a comparative analysis.
Previous comparative analyses fitting
heavy-tailed distributions
(distributions with higher probabilities of extreme events
than the normal distribution)
to time series
have not accounted for an underlying population dynamics model
\citep{keitt1998, allen2001, halley2002, segura2013}.
Alternatively, most examples of population die offs come from identifying
sudden changes in abundance that exceed a threshold of decline
\citep[e.g.][]{young1994, gerber2001, reed2003, fey2015}.
But this approach gives no indication of relative event frequency
and again does not distinguish events from
expected dynamics.

@@

We fit population dynamics models in which the process noise is
drawn from a Student-$t$ distribution. By estimating the degrees of freedom
parameter, \(\nu\), we can estimate the degree to which the process deviations
have heavy tails and are therefore diagnostic of black-swan events (Fig.~1).
Lower values of \(\nu\) result in heavy-tailed distributions.
% For example, at \(\nu = 2\), an event
% more than 4 standard deviations below the mean is expected
For example, an event four standard deviations below the mean
is expected to happen once every 35 years with a $t$ distribution at \(\nu = 2\)
compared to once every 32,000 years with a normal distribution.
% more than four standard deviations below the mean is almost 70,000 times higher
% than the probability of drawing such a value from a normal distribution
% (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)).
As \(\nu\) approaches infinity, the t
distribution approaches the normal distribution (Fig.~1).

@@

One alternative, the generalized extreme value (GEV) distribution,
represents the limit distribution of a series of maxima or minima.
Although the GEV is well-suited for physical ecological variables
\citep[e.g.\ sedimentation rates or wind speeds;][]{gaines1993, denny2009},
the GEV requires maxima per time block
or peaks beyond a chosen threshold.
Therefore, the GEV requires longer time series then typically available
for wild populations.
Furthermore, the GEV approach cannot be easily integrated into population dynamics models.

% This could represent the effect of unmodeled spatial heterogeneity,
% nonstationary environmental processes, or
% varying responses to temperature among individuals, for example.

@@

We selected abundance time series from the Global Population Dynamics
Database \citep[GPDD,][]{gpdd2010}, which contains nearly 5,000 time series of
abundance from \(\sim\) 1,000 species and \(\sim\) 100 taxonomic orders. We
filtered the data (Supporting Information) to remove populations from less
reliable data sources, and those without sufficient data for our models, and
then interpolated remaining missing values (sensu \citet{brook2006a}).
Our interpolation affected only \(\sim\)
\interpPointsPerc \% of the final data points (Table~S1) and none of the data
points that were later considered black-swan events. Our final dataset
contained \NPops\ populations across \NOrders\ taxonomic orders and seven
taxonomic classes, with a median of \medianTimeSteps\ time steps (range
\minTimeSteps--\maxTimeSteps) (Table~S1).

@@

\subsubsection{Density independence} We fit a simplified version of the
Gompertz model in which the density dependence parameter \(b\) was fixed at
\(1\) (density independent). This is equivalent to fitting a random walk model
(with drift) to the \(\ln\) abundances or assuming the growth rates are drawn
from a stationary distribution. The model was as follows:
\begin{align}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
\end{align}
We fit this model for three reasons: (1) it is computationally simpler and so
provides a check that our more complicated full Gompertz model was obtaining
reasonable estimates of \(\nu\), (2) it provides a test of whether density
dependence was systematically affecting our perception of heavy tails, (3) it
matches how some previous authors have modelled heavy tails without accounting
for density dependence \citep{segura2013}.

\subsubsection{Observation error} Observation error can bias parameter
estimates \citep{knape2012} and is known to affect the ability to detect extreme
events \citep{ward2007}. In our main analysis, we fit a model that ignored
observation error. One way to account for observation error would be to fit a
full state-space model that simultaneously estimates the magnitude of process
noise and observation error. However, simultaneously estimating observation and
process noise is a challenging problem (e.g.\ because the observation and
process noise parameters tend to negatively covary in model fitting) and is
known to sometimes result in identifiability issues with the Gompertz
population model \citep{knape2008}. Furthermore, our model was applied to
hundreds of time series, often of short length (as few as \minTimeSteps\ time
steps) and our model estimates an additional parameter---the shape of the
process deviation tails---potentially making identifiability and computational
issues even greater. Therefore, we considered a version of the base Gompertz
model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
represents the standard deviation of observation error (on a log scale), and
\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
simulation analyses \citep[e.g.][]{valpine2002, thorson2014b}.

\subsubsection{Ricker-logistic} We also fit a Ricker-logistic model:
\begin{align}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma),
\end{align}
where \(r_\mathrm{max}\) represents the theoretical maximum population growth
rate that is obtained when \(N_t = 0\). The parameter \(K\) represents the
carrying capacity and, as before, \(x_t\) represents the \(\ln\) transformed
abundance at time \(t\). The Ricker-logistic model assumes a linear decrease in
population growth rate with increases in abundance. In contrast, the Gompertz
model assumes a linear decrease in population growth rate with increases in
\textit{ln} abundance.

@@

We found that we could consistently and precisely recover median posterior
estimates of \(\nu\) near the true value of \(\nu\) with large samples (\(\ge
200\)) (Fig.~S\ref{fig:sim-nu} upper panels). At smaller samples we could
usually qualitatively distinguish heavy from not-heavy tails, but the model
tended to underestimate how heavy the tails were. At the same time, at smaller
sample sizes, the model tended to overestimate how large the scale parameter
was (Fig.~S\ref{fig:sim-nu} lower panels).

@@

\subsection{Skewed Student-$t$ forecasts} To evaluate the apparent skewness
of heavy-tailed process noise we fit Gompertz models with skewed Student-$t$
distributed process noise \citep{fernandez1998}. This distribution adds one
parameter to the Student-$t$ distribution, \(\gamma\), which controls the
skewness. The distribution is symmetrical if \(\gamma = 1\), left skewed if \(0
< \gamma < 1\), and right skewed if \(1 < \gamma < \infty\). We placed a weakly
informative prior of \(\mathrm{Cauchy}(0, 2.5)\) on \(\log \gamma\). We
aggregated 5,000 randomly selected posterior samples from the \(\gamma\)
parameter of each model at three levels of evidence for heavy-tails:
\emph{heavy tailed}, median \(\nu < 10\); \emph{slightly heavy tailed}, \(10 \le
\mathrm{median}\, \nu \le 70\); and \emph{normal tailed}, median \(\nu > 70\)
(Fig.~4).

@@

Given our results, we suggest modelers should routinely consider representing population dynamic
process noise with a heavy-tailed, and possibly downward-skewed, distribution,
especially when making forecasts to evaluate risk.
But what heavy-tailed distribution should one pick
and how heavy tailed should it be?
% nearly any reason for why a population modeler has incorrectly specified the exact and full process generating the abundance times could generate a heavy tailed process noise.
This is an open research question.
The exact shape of the tails has proven important
in the field of dispersal biology \citep{kot1996, clark1999}.
Answers for the field of population dynamics
might come from simulation analyses exploring
the implications of different tail shapes and
from comparing the relative fit of heavy-tailed
population models to time series
with suspected mass mortality events
\citep[e.g.~the $t$ distribution compared to a mixture
of two normal distributions;][]{ward2007}.
Nevertheless, our proposed $t$ (or the skew-$t$) distribution
is a simple and elegant solution that
can be easily substituted for the normal distribution
Furthermore, if the degrees of freedom parameter, $\nu$,
is fixed
it requires estimating no more parameters
than the normal distribution.
Barring a better estimate,
we suggest using $\nu = 4$; and
a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a)
if a modeler wishes to allow for downward heavy-tailed events.
These are midrange values from the populations identified as heavy tailed here.
Above $\nu  =  10$ the $t$ distribution closely resembles
the normal distribution,
and $\nu  = 4$ allows for an event
three standard deviations below the mean to happen
once every 50 years instead of once every 740 years.
Adopting heavy-tailed process noise by default follows the intuition
that our models are usually reasonable but occasionally very wrong.

@@

\textbf{Figure~1} Illustration of population dynamic models that allow for heavy
tails. (a to b) Probability density for the Student-$t$
distribution with scale parameter of $1$ and different values of \(\nu\). Small
values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the
distribution approaches the normal distribution. (c to e)
Simulated population dynamics from a Gompertz model with process noise drawn
from Student-$t$ distributions with three values of \(\nu\). Coloured dots in
panels c and d represent jumps with less than a 1 in 1000
chance of occurring in a normal distribution. (f to h)
Estimates of \(\nu\) from models fit to the times series in panels c to
e. Shown are posterior samples (histograms), median and interquartile
range of the posterior (IQR, dots and line segments), and the exponential prior
on \(\nu\) (dashed lines). Colour shading behind panels f to h
illustrates the approximate region of heavy tails.

@@

\textbf{Figure~2} Estimates of population dynamic heavy-tailedness for
populations of birds, mammals, insects, and fishes. (a to d)
Small values of \(\nu\) suggest heavy-tailed black-swan dynamics. Vertical
points and line segments represent posterior medians and 50\% / 90\% credible
intervals for individual populations. Inset plots show probability that \(\nu
< 10\) for populations arranged by taxonomic order and sorted by decreasing
mean Pr(\(\nu < 10\)). Taxonomic orders with three or fewer populations in
panel a are omitted for space. Red to yellow points highlight populations
with a high to moderately high probability of heavy-tailed dynamics.

@@

\textbf{Figure~3} Standardized probabilities and covariates of heavy-tailed
dynamics. (a) Taxonomic-order-level posterior densities of Pr(\(\nu
< 10\)) after accounting for time-series length in a hierarchical model. Dotted
vertical line in a indicates the median expected Pr(\(\nu < 10\)) from the
prior distribution. Colour shading refers to taxonomic class (yellow: fishes,
green: insects, purple: birds, and red: mammals). Estimates are standardized at
the geometric mean of time series length across all the data (approximately 27
time steps) and shown for orders with \(\ge 5\) populations. (b)
Posterior densities for potential covariates of Pr(\(\nu < 10\)). In both
panels, short vertical line segments within the density polygons indicate
medians.

@@

\textbf{Figure~4} Heavy-tailed process noise tends to be downward skewed and
ignoring this can underestimate risk. (a) Illustrations of Student-$t$
distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)).
(b) Posterior density of the skewness parameters aggregated across
populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly
heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations
(\(\hat{\nu} > 70\)). Approximate mid-values from b are illustrated in
panel a. (c) Example time series of heavy-tailed populations with
different levels of skewness. Red dots highlight likely heavy-tailed events.
Forecasts (grey regions) show median (solid lines) and lower 99\% credible
intervals (dotted lines) of abundance. Black and red lines indicate forecasts
from Gompertz models with lognormal and skew-$t$ process noise, respectively.
